Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-17 AppendixI.Rmd
Appendix, redundancy

**Figure A2**: Trajectories of seed mass classes proportions over 30 years after disturbance, corresponding to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. No gap filling process was applied in this case. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3 \@ref(fig:FigSeedMassClasses).

**Figure A3**: Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities **(a)** Richness, **(b)** Shannon and **(c)** Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3 \@ref(fig:DivTaxo).

**Figure A4**: Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes. \@ref(fig:DivFun).

**Figure A5**: Summarizing plots of the functional PCA analysis defining communities functional space. **(a)** Plot of the samples from the BRIDGE functional database in the two dimensional main plan. **(b)** bow plot of the PCA eignevalues in percentage. **(c)** Plot of the 7 functional traits sed in the PCA analysis (Leaf and stem traits, and maximum species height). \@ref(fig:PCA_graphs).

**Figure A6**: Trajectories of the functional redundancy over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3. \@ref(fig:RedFun).

@@

# **Figure A2**: Proportion of seed mass classes along time

@@

# **Figure A3**: Richness, Shannon and Simpson taxonomic diversity trajectories

@@

# **Figure A4**: Rao diversity trajectories with all plots display

@@

# **Figure A5**: Functional PCA

@@

# **Figure A6**: Overall functional redundancy

@@

par(mar=c(5,4,5,2))
RedundancyPlot(RedundancyTraj)
mtext("Functional Redundancy",side=3,adj=0,line=1.5)
mtext("Restricted to initial functional space",side=3,adj=0,cex=0.8,line=0.5)
mtext("Years since disturbance",side=1,line=2.2,adj=1)

>>>>

**Figure A2**: Scheme of the redundancy measurement process

**Figure A3**: Trajectories of seed mass classes proportions over 30 years after disturbance, corresponding to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. No gap filling process was applied in this case. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3 \@ref(fig:FigSeedMassClasses).

**Figure A4**: Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities **(a)** Richness, **(b)** Shannon and **(c)** Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3 \@ref(fig:DivTaxo).

**Figure A5**: Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes. \@ref(fig:DivFun).

**Figure A6**: Summarizing plots of the functional PCA analysis defining communities functional space. **(a)** Plot of the samples from the BRIDGE functional database in the two dimensional main plan. **(b)** bow plot of the PCA eignevalues in percentage. **(c)** Plot of the 7 functional traits sed in the PCA analysis (Leaf and stem traits, and maximum species height). \@ref(fig:PCA_graphs).

**Figure A7**: Trajectories of the functional redundancy over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3. \@ref(fig:RedFun).

@@

# **Figure A2**: Redundancy metric

```{r RedMM,out.width='100%',fig.cap="Scheme of the redundancy measurement process",echo=F, fig.env="figure*"}
knitr::include_graphics("ExternalFig/RedundancyMetric.png")
```

# **Figure A3**: Proportion of seed mass classes along time

@@

# **Figure A4**: Richness, Shannon and Simpson taxonomic diversity trajectories

@@

# **Figure A5**: Rao diversity trajectories with all plots display

@@

# **Figure A6**: Functional PCA

@@

# **Figure A7**: Overall functional redundancy

@@

par(mar=c(5,4,5,2))
RedundancyPlot(RedundancyTraj)
mtext("Functional Redundancy",side=3,adj=0,line=0.5)
mtext("Years since disturbance",side=1,line=2.2,adj=1)

### 2018-05-17 AppendixI.tex
Appendix, redundancy

\textbf{Figure A2}: Trajectories of seed mass classes proportions over
30 years after disturbance, corresponding to the median (solid line) and
0.025 and 0.975 percentile (gray envelope) observed after 50 iteration
of the taxonomic uncertainty propagation. No gap filling process was
applied in this case. Initial treatments are represented by solid lines
colors with green for control, blue for T1,orange for T2 and red for T3
\ref{fig:FigSeedMassClasses}.

\textbf{Figure A3}: Trajectories of the difference to the 1989
inventories (5 years after disturbance) over 30 years after disturbance
of plots communities \textbf{(a)} Richness, \textbf{(b)} Shannon and
\textbf{(c)} Simpson diversities. Trajectories correspond to the median
(solid line) and 0.025 and 0.975 percentile (gray envelope) observed
after 50 iteration of the taxonomic uncertainty propagation. Initial
treatments are represented by solid lines colors with green for control,
blue for T1,orange for T2 and red for T3 \ref{fig:DivTaxo}.

\textbf{Figure A4}: Trajectories of the Rao functional diversity over 30
years after disturbance. Trajectories correspond to the median (solid
line) and 0.025 and 0.975 percentile (gray envelope) observed after 50
iteration of the taxonomic uncertainty propagation. Initial treatments
are represented by solid lines colorswith green for control, blue for
T1,orange for T2 and red for T3.and the missing trait value filling
processes. \ref{fig:DivFun}.

\textbf{Figure A5}: Summarizing plots of the functional PCA analysis
defining communities functional space. \textbf{(a)} Plot of the samples
from the BRIDGE functional database in the two dimensional main plan.
\textbf{(b)} bow plot of the PCA eignevalues in percentage. \textbf{(c)}
Plot of the 7 functional traits sed in the PCA analysis (Leaf and stem
traits, and maximum species height). @ref(fig:PCA\_graphs).

\textbf{Figure A6}: Trajectories of the functional redundancy over 30
years after disturbance. Trajectories correspond to the median (solid
line) and 0.025 and 0.975 percentile (gray envelope) observed after 50
iteration of the taxonomic uncertainty propagation and the missing trait
value filling processes. Initial treatments are represented by solid
lines colors with green for control, blue for T1,orange for T2 and red
for T3. \ref{fig:RedFun}.

@@

\section{\texorpdfstring{\textbf{Figure A2}: Proportion of seed mass
classes along
time}{Figure A2: Proportion of seed mass classes along time}}\label{figure-a2-proportion-of-seed-mass-classes-along-time}

@@

\section{\texorpdfstring{\textbf{Figure A3}: Richness, Shannon and
Simpson taxonomic diversity
trajectories}{Figure A3: Richness, Shannon and Simpson taxonomic diversity trajectories}}\label{figure-a3-richness-shannon-and-simpson-taxonomic-diversity-trajectories}

@@

\section{\texorpdfstring{\textbf{Figure A4}: Rao diversity trajectories
with all plots
display}{Figure A4: Rao diversity trajectories with all plots display}}\label{figure-a4-rao-diversity-trajectories-with-all-plots-display}

@@

\section{\texorpdfstring{\textbf{Figure A5}: Functional
PCA}{Figure A5: Functional PCA}}\label{figure-a5-functional-pca}

@@

\section{\texorpdfstring{\textbf{Figure A6}: Overall functional
redundancy}{Figure A6: Overall functional redundancy}}\label{figure-a6-overall-functional-redundancy}

>>>>

\textbf{Figure A2}: Scheme of the redundancy measurement process

\textbf{Figure A3}: Trajectories of seed mass classes proportions over
30 years after disturbance, corresponding to the median (solid line) and
0.025 and 0.975 percentile (gray envelope) observed after 50 iteration
of the taxonomic uncertainty propagation. No gap filling process was
applied in this case. Initial treatments are represented by solid lines
colors with green for control, blue for T1,orange for T2 and red for T3
\ref{fig:FigSeedMassClasses}.

\textbf{Figure A4}: Trajectories of the difference to the 1989
inventories (5 years after disturbance) over 30 years after disturbance
of plots communities \textbf{(a)} Richness, \textbf{(b)} Shannon and
\textbf{(c)} Simpson diversities. Trajectories correspond to the median
(solid line) and 0.025 and 0.975 percentile (gray envelope) observed
after 50 iteration of the taxonomic uncertainty propagation. Initial
treatments are represented by solid lines colors with green for control,
blue for T1,orange for T2 and red for T3 \ref{fig:DivTaxo}.

\textbf{Figure A5}: Trajectories of the Rao functional diversity over 30
years after disturbance. Trajectories correspond to the median (solid
line) and 0.025 and 0.975 percentile (gray envelope) observed after 50
iteration of the taxonomic uncertainty propagation. Initial treatments
are represented by solid lines colorswith green for control, blue for
T1,orange for T2 and red for T3.and the missing trait value filling
processes. \ref{fig:DivFun}.

\textbf{Figure A6}: Summarizing plots of the functional PCA analysis
defining communities functional space. \textbf{(a)} Plot of the samples
from the BRIDGE functional database in the two dimensional main plan.
\textbf{(b)} bow plot of the PCA eignevalues in percentage. \textbf{(c)}
Plot of the 7 functional traits sed in the PCA analysis (Leaf and stem
traits, and maximum species height). @ref(fig:PCA\_graphs).

\textbf{Figure A7}: Trajectories of the functional redundancy over 30
years after disturbance. Trajectories correspond to the median (solid
line) and 0.025 and 0.975 percentile (gray envelope) observed after 50
iteration of the taxonomic uncertainty propagation and the missing trait
value filling processes. Initial treatments are represented by solid
lines colors with green for control, blue for T1,orange for T2 and red
for T3. \ref{fig:RedFun}.

@@

\section{\texorpdfstring{\textbf{Figure A2}: Redundancy
metric}{Figure A2: Redundancy metric}}\label{figure-a2-redundancy-metric}

{\centering \includegraphics[width=1\linewidth]{ExternalFig/RedundancyMetric}

\caption{Scheme of the redundancy measurement process}\label{fig:RedMM}
\end{figure*}

\section{\texorpdfstring{\textbf{Figure A3}: Proportion of seed mass
classes along
time}{Figure A3: Proportion of seed mass classes along time}}\label{figure-a3-proportion-of-seed-mass-classes-along-time}

@@

\section{\texorpdfstring{\textbf{Figure A4}: Richness, Shannon and
Simpson taxonomic diversity
trajectories}{Figure A4: Richness, Shannon and Simpson taxonomic diversity trajectories}}\label{figure-a4-richness-shannon-and-simpson-taxonomic-diversity-trajectories}

@@

\section{\texorpdfstring{\textbf{Figure A5}: Rao diversity trajectories
with all plots
display}{Figure A5: Rao diversity trajectories with all plots display}}\label{figure-a5-rao-diversity-trajectories-with-all-plots-display}

@@

\section{\texorpdfstring{\textbf{Figure A6}: Functional
PCA}{Figure A6: Functional PCA}}\label{figure-a6-functional-pca}

@@

\section{\texorpdfstring{\textbf{Figure A7}: Overall functional
redundancy}{Figure A7: Overall functional redundancy}}\label{figure-a7-overall-functional-redundancy}

### 2018-05-17 WholePlotTrajectories.Rmd
Appendix, redundancy

The functional redundancy was measured as the overlap among species in communities' functional space [@Carmona2016].
The samples of the trait database were first located with a PCA analysis in a two-dimensional functional space summarizing the 8 functional traits considered.
We choose to summarize species functional space through a PCA because functional traits considered were already somehow redundant and allows not to overweight the role of one trait.
Then, multivariate kernel density estimator associated with individual trees were summed to give the distribution of traits probabilites of each species.
For each community the trait probability distributions of corresponding species were combined and weighted by species abundance. 
Eventually communities functional redundancy was measured as the sum of overlap between species weighted functional density.
Communities functional redundancy is expressed as the average number of species that could be removed from the community without reducing the functional space.

>>>>

The functional redundancy is measured as the overlap among species in communities' functional space [@Carmona2016]. First, the samples of the trait database are mapped in a 2-dimensional plan with a PCA analysis. Then, multivariate kernel density estimator associated with individual trees were give species traits probability distribution (TDP).
TDP, weighted by species abundance, are summed to give communities TDP: the functional redundancy is the sum of TDP overlaps,  corresponding to the average number of species that could be removed from without reducing the functional space (see appendix I for a more comprehensive sheme).

### 2018-05-17 WholePlotTrajectories.tex
Appendix, redundancy

In tropical forest, ecological communities are contantly re-shaped by
the natural disturbance events that changes the abiotic environment,
through the fluxes of light, heat and water, and the biotic interaction
and competitive pressure \citep{Goulamoussene2017}. The cornerstone of
tropical forests ecology is then to understand the underlyings of
ecosystem response to disturbance, its mechanisms and drivers
\citep{White2001, Chazdon2003a}. For now, this has been largely studied
through structural parameters, rapid and convenient to measure, as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. These structural parameters have
then been sucessfully modeled and allowed to assess the maintenance of
ecosystems processes and services \citep{Denslow2000, Blanc2009}.
However the response of tree species diversity remains unclear, also it
is determinant of ecosystems productivity, stability and functioning
\citep[\citet{Liang2016}]{Tilman2014} and would be most probably
impacted by post-disturbance environmental changes
\citep{Baraloto2012a}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Validations of the IDH in
the long term, however, remain scarce and mainly rely on the analysis of
the rough species richness that gives limited or misleading information
on forests recovery and functioning \citep{Martin2015, Chaudhary2016}.
More releveant monitoring would encompass communities composition, that
is crucial for conservation issues, and diversity profiles, that reveals
ecological rules underlying communities' structure
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, the
functional approach accounting for differenes species biological
attributes would be insightful as it reveals the species role in the
post-disturbance trajectory of ecosystems functioning
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect major functional traits related to species ecology and
performance were largely adopted through an integrative framework
\citep{Diaz2005, Villeger2008a}. The functional trait-based approach,
for example, highlighted in tropical rainforests the environmental
filters fostering disturbance resistant species with rapid growth and
efficient resources acquisition \citep{Molino2001, Haddad2008}. In
disturbed tropical rainforests, this framework translates into the
fostering of disturbance resistant species with rapid growth and
efficient resources acquisition \citep{Molino2001, Haddad2008}. It
results in a shift from ``conservative'' slow-growing species dealing
with scarce resources that dominate before disturbance, to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{TerSteege2001, Reich2014, Herault2011}. It was
mirrored by shifts in key functional traits related to resource
acquisition (leaf area, density and chlorophyll content, and stem
specific gravity and bark thickness), tree growth and reproduction life
history traits (seed mass and maximum height)
\citep{Wright2004, Westoby2006a, Chave2009b}.

A proper monitoring of communities response should therefore encompass
taxonomic and functional diversity and composition measures to test the
validity of the IDH in the long term for tropical forests, and clarify
the resilience of communities evenness, composition, and also
functioning. The trajectories followed by all these facets would
highlight the role of deterministic processes, like competitive
exclusion or abiotic selection, and the communities' convergence
maintaining intrinsinc differences in diversity and composition,which is
as much insgights for future adaptive conservation strategies
\citep{Adler2007}.

Here we investigated over 30 years the response of 75 ha of forests
plots set up on a gradient of disturbance intensity, from 10 to 60\% of
ecosystem biomass removed. We made use of a large functional traits
database browsing major leaf, stem and seed traits and species maximum
height to draw the trajectories over time of communities taxonomic and
functional composition and diversity. Specifically, we (i) tested the
validity of the IDH in the long term for tropical hyperdiverse forest
and highlighted the ecological rules shaping their response to
disturbance, (ii) clarified the different facets of communities
resilience in terms of communities composition, diversity and
functioning \emph{????(iii) questioned the completeness of communities
recovery given the altered functional redundancy.}

@@

The functional redundancy was measured as the overlap among species in
communities' functional space \citep{Carmona2016}. The samples of the
trait database were first located with a PCA analysis in a
two-dimensional functional space summarizing the 8 functional traits
considered. We choose to summarize species functional space through a
PCA because functional traits considered were already somehow redundant
and allows not to overweight the role of one trait. Then, multivariate
kernel density estimator associated with individual trees were summed to
give the distribution of traits probabilites of each species. For each
community the trait probability distributions of corresponding species
were combined and weighted by species abundance. Eventually communities
functional redundancy was measured as the sum of overlap between species
weighted functional density. Communities functional redundancy is
expressed as the average number of species that could be removed from
the community without reducing the functional space.

@@

Although both communities taxonomic and functional characteristics
proved resilient and followed similar humped-back trajectories, the
taxonomic recovery systematically lagged behind the corresponding
functional dynamics. Such delay between functional and taxonomic
dynamics has already been observed for grasslands
\citep{Tilman1997, Mouillot2011} and more recently for tropical forests
\citep{Lohbeck2015, Guariguata2001}. According to the ``vegetation
quantity effect'' \citep{Grime1998}, functional trajectories rely on the
pool of dominant species which diversity and evenness were enhanced
after disturbance and which rapidly restored the functional diversity
and composition. However, infrequent species that would further lower
the functional diversity and communities evenness still missed to the
recovery of taxonomic characteristics. Taxonomic recovery then
mechanically lagged behind, all the more so that unrecovered species are
functionally redundant and would probably undergo competitive pressures.

@@

The monitoring of disturbed forest communities confirmed a limited or
negative impact of intense disturbance on species richness, as observed
on several post logging surveys \citep{Cannon1998, Baraloto2012a}, while
. Thirty years after disturbance, the genus richness was restored to
initial plot values after high disturbance intensity while it
substantially increased after low disturbance intensity, reaching a gain
of almost 12 genera for some plots.

Both richness and evenness followed asymptotic trajectories after
disturbance, sharply increasing for 15 years before stabilizing at
higher values than those of control plots. An increase of communities
evenness stems from a higher homogeneity of species distribution. After
disturbance, communities are made of the old, pre-disturbance survivors
and the newly recruited trees: changes in composition and abundance
distribution are to be found in the recruitment processes and in the
pre-disturbance survivors mortality. Because the composition of old
survivors proved to mirror the initial community \citep{Herault2018}, a
specific turnover was expected among recruited trees with enhanced
growth and survival of previously infrequent species. Indeed, the
increase in taxonomic diversity was accompanied by an increase of
taxonomic dissimilarity with plots initial state and a functional shift
towards resource-acquisitive strategies (sharp increase in the SLA, leaf
thickness and bark thickness and decrease in wood density, leaf
toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Disturbance then causes a
reorganization of the typical high dominance structure of hyperdiverse
mature forests after disturbance, benefiting to pioneers and light
demanding species. Likely, the changes in abiotic environment and
competitive pressure favored pioneers which outcompete others in non
limiting resources but are excluded in mature forests by long-lived,
more resistant and shade tolerant species.

As stated by the IDH, communities dynamics after disturbance relied on
species functional strategy and corresponding ability to fill the
environmental niches made available by disturbance. Recruited species
then mixed with pre-disturbance ones, from which they differed, and
constituted a community all the more diversified that the disturbance
was intense \citep{Molino2001}.

\subsection{Functional redundancy of disturbed
ecosystems}\label{functional-redundancy-of-disturbed-ecosystems}

The functional redundancy, the functional overlap between species that
is typical of the huge biodiversity of tropical forests
\citep{Bellwood2006}, was then not restored 30 years after disturbance
and this is major to consider as it defines forests' resilience
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Besides the long-term alteration of functional redundancy, there was
probably persistent compositional changes favoring disturbance resistant
species \citep{Haddad2008}, lianas or epiphytes \citep{Martin2013} and
environmental changes, like in the soils nutrient cycling and compaction
\citep{Olander2005}. These persistent changes highly question forest's
resilience \citep{Chazdon2003a}. New conditions would not only be longer
lasting but self-maintained as tied to disturbance regime
\citep{Burslem2000}. Specificaly, this would impair species contingent
to undisturbed forests, threatening their maintenance, and run the risk
to loose cornerstone species and trigger unexpected ecological
consequences \citep{Jones1994, Diaz2005, Gardner2007}.

@@

Our study showed the significant impact of disturbance on tropical
forests communities and validated the consistency of the IDH in the long
term. It revealed the contrasting response of communities taxonomic and
functional characteristics, with persisting impacts on the species
abundance distibution while the functional diversity and dominant
functional strategies were restored. Communities recovery therefore
remained unachieved but consistent for the range of disturbance studied
here. The length of the recovery, however, severly questioned the
sustainability of intense selective logging and advocated felling cycle
much longer than 30 years \citep{Gourlet-Fleury2005}.

>>>>

In tropical forest, ecological communities are contantly re-shaped by
the natural disturbance events changing both abiotic environment,
through the fluxes of light, heat and water, and biotic interactions and
competitive pressure \citep{Goulamoussene2017}. The cornerstone of
tropical forests ecology is to understand the underlyings of ecosystem
response to disturbance, its mechanisms and drivers
\citep{White2001, Chazdon2003a}. For now, this has been largely studied
through structural parameters, rapid and convenient to measure, as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Consistent models assessed these
structural parameters and highlighted the maintenance of ecosystems
processes and services \citep{Denslow2000, Blanc2009}. The response of
tree diversity, however, remains unclear albeit it determines ecosystems
productivity, stability and functioning
\citep[\citet{Liang2016}]{Tilman2014} and would be most probably
impacted by post-disturbance environmental changes
\citep{Baraloto2012a}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Validations of the IDH in
the long term, however, remain scarce and mainly rely on the analysis of
the rough species richness that gives limited or misleading information
on forests recovery and functioning \citep{Martin2015, Chaudhary2016}.
More releveant monitoring would encompass communities composition, that
is crucial for conservation issues, and diversity profiles, that reveals
ecological rules underlying communities' structure
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, account
for species biological attributes and role the ecosystem through a
functional approach assesses communities functioning and corresponding
mechanisms
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect major functional traits related to species ecology and
performance were largely adopted \citep{Diaz2005, Villeger2008a}
highlighting, for example, the environmental filters fostering
disturbance resistant species with rapid growth and efficient resources
acquisition in tropical rainforests \citep{Molino2001, Haddad2008}. In
disturbed tropical rainforests, this framework translates into the
fostering of disturbance resistant species with rapid growth and
efficient resources acquisition \citep{Molino2001, Haddad2008}. It
results in a shift from ``conservative'' slow-growing species dealing
with scarce resources that dominate before disturbance, to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{TerSteege2001, Reich2014, Herault2011}. It was
mirrored by shifts in key functional traits related to resource
acquisition (leaf area, density and chlorophyll content, and stem
specific gravity and bark thickness), tree growth and reproduction life
history traits (seed mass and maximum height)
\citep{Wright2004, Westoby2006a, Chave2009b}.

Functional and taxonomic trajectories relate to distinct characteristics
and community processes, and might therefore be decoupled as already
observed in tropical forests \citep{Lohbeck2015, Guariguata2001}. The
combination of taxonomic and functional characteristics is translated by
the functional redundancy, measuring the amount of species sharing same
trait values \citep{Carmona2016}. Functional redundancy is a major
determinant of communities resilience as high redundancy, like in highly
diverse tropical forests \citep{Bellwood2006}, mitigates the impacts of
species removal on ecosystem functioning
\citep{Trenbath1999, Elmqvist2003, Diaz2005}. It is therefore a key
element as a link between taxonomic and functional diversities and as a
determinant of communities resilience.

Grasping all facets of communities response to disturbance then requires
to disentangle the taxonomic and functional trajectories in diversity
and composition. These trajectories would highlight communities
resilience and the drivers of their response to disturbance, clarifying
the tenants of the IDH in the long term for tropical forests, that are
as much insights for future adaptive conservation strategies
\citep{Adler2007}. Here we monitored over 30 years the response of 75 ha
of forests plots set up on a gradient of disturbance intensity, from 10
to 60\% of ecosystem biomass removed. We made use of a large functional
traits database browsing major leaf, stem and seed traits and species
maximum height to draw the trajectories over time of communities
taxonomic and functional composition and diversity. Specifically, we (i)
highlighted the determinant of communities functoinal and taxonomic
response to disturbance, (ii) tested the validity of the IDH in the long
term for tropical forest and (iii) questioned the completeness of
communities recovery regarding their functional redundancy.

@@

The functional redundancy is measured as the overlap among species in
communities' functional space \citep{Carmona2016}. First, the samples of
the trait database are mapped in a 2-dimensional plan with a PCA
analysis. Then, multivariate kernel density estimator associated with
individual trees were give species traits probability distribution
(TDP). TDP, weighted by species abundance, are summed to give
communities TDP: the functional redundancy is the sum of TDP overlaps,
corresponding to the average number of species that could be removed
from without reducing the functional space (see appendix I for a more
comprehensive sheme).

@@

Although both communities taxonomic and functional characteristics
proved resilient and followed similar humped-back trajectories, the
taxonomic recovery systematically lagged behind the corresponding
functional dynamics. Such delay between functional and taxonomic
dynamics has already been observed for grasslands
\citep{Tilman1997, Mouillot2011} and more recently for tropical forests
\citep{Lohbeck2015, Guariguata2001}. According to the ``vegetation
quantity effect'' \citep{Grime1998} functional trajectories rely on the
pool of dominant species, which diversity and evenness were enhanced
after disturbance, and which rapidly restored the functional diversity
and composition. However, communities evenness remained high so
infrequent species still missed to the taxonomic recovery that
mechanically lagged behind, all the more so that unrecovered species
would be functionally redundant and probably undergo competition
processes.

@@

Validating the IDH, communities trajectories confirmed the diversity
increase after disturbance through an enhanced growth of otherwise less
favored species. This was however completely clear regarding the
functional diversity but much more blurry for communities taxonomic
characteristics.

The taxonomic richness was weakly or negatively impacted by intense
disturbance, as observed on several post logging surveys
\citep{Cannon1998, Baraloto2012a}, while it substantially increased
after low intensity disturbance. Disturbance also enhanced taxonomic
evenness, but its intensity was only weakly and lately correlated to
communities eveness (\emph{i.e.} Simpson diversity). Contrastingly,
disturbance intensity consistently predicted the significant increase of
communities functional diversity for 30 years. Disturbance impacts
involve specific turnover within communities, either among
pre-disturbance survivors or among newly recruited trees. As the
composition of old-growth survivors proved to mirror initial communities
\citep{Herault2018}, disturbance would impact trees recruitment through
the enhanced growth and survival of previously infrequent species and
functional characteristics. Consistently, disturbance resulted in
increasing taxonomic dissimilarity compared to pre-disturbance
communities and functional shifts towards resource-acquisitive
strategies (sharp increase in the SLA, leaf thickness and bark thickness
and decrease in wood density, leaf toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Disturbance then caused a
reorganization of the typical high dominance structure of hyperdiverse
mature forests after disturbance, benefiting to pioneers and light
demanding species. Likely, the changes in abiotic environment and
competitive pressure favored pioneers which outcompete others in non
limiting resources but are excluded in mature forests by long-lived,
more resistant and shade tolerant species. Therefore, consistently with
communities dynamics after disturbance relied on species functional
strategy and corresponding ability to fill the environmental niches made
available by disturbance. Recruited species then mixed with
pre-disturbance ones, from which they differed, and constituted a
community all the more diversified that the disturbance was intense
\citep{Molino2001}.

\subsection{The functional redundancu and the resilience of
communities}\label{the-functional-redundancu-and-the-resilience-of-communities}

Both the lag between taxonomic and functional recovery and the middling
consistency of the IDH regarding communities taxonomic diversity entail
a central role of functional redundancy. Functional redundancy,
determinant of communities resilience
\citep{Trenbath1999, Elmqvist2003, Diaz2005}, seemed not to have
recovered 30 years after disturbance.

Although disturbance would be mitigated by communities functional
redundancy, we expect the redundancy itself to be reduced or
re-organized within the functional space. Globally, disturbed
communities did not follow consistent trajectories after disturbance and
30 years after disturbance, irrespective of the intensity, disturbed
communities displayed lower, similar or higher functional redundancy.
The redundancy within the functional space of the initial community
however clearly followed humped-back trajectories which decrease was
determined by the disturbance intensity and which return to initial
state was not observed for any disturbed community. Not only did the
functional redundancy and hence the resilience of pre-disturbance
community decrease, but also did the ones of disturbance-specific
communities increase. It means higher chances to have long lasting or
self-maintained compositional changes towards disturbance resistant
species, lianas or epiphytes \citep{Haddad2008, Burslem2000, Martin2013}
and thus highly question forest's resilience \citep{Chazdon2003a}.
Specificaly, this would impair species contingent to undisturbed
forests, threatening their maintenance, and run the risk to loose
cornerstone species and trigger unexpected ecological consequences
\citep{Jones1994, Diaz2005, Gardner2007}.

@@

our study disentangled tropical forests functional and taxonomic
decoupled response to disturbance, with a rapid recovery of communities
functioning but slower and more variable taxonomic dynamics.
Consistently with the IDH functional trajectories were constrained by
environmental pressures favoring pioneers and light-demanding species
after disturbance, but taxonomic dynamics were more stochastic as they
involved infrequent speciesfunctoinally redundant with the dominants and
thus under competitive pressure. While communities functioning rapidly
recovered, taxonomic trajectories, although consistent and converging
towards full recovery, were much slower and entailed persisting
alteration of communities functional redundancy. the trajectories
therefore suggested a potential sustainability of tropical forests in
the face of quite intense disturbance but only when followed by long
recovery periods, longer than 30 years \citep{Gourlet-Fleury2005}. This
study besides highlighted the central role of recruitment processes
underlying the IDH mechanisms and holding the whole community response
to disturbance.
